Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-11 WholePlotTrajectories.Rmd
Redundancy plot

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   #Pour avoir plusieurs lignes qui suivent. Commencer à 2 caractères
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

@@

In tropical forest, ecological communities are contantly re-shaped by the natural disturbance events that changes the abiotic environment, through the fluxes of light, heat and water, and the biotic interaction and competitive pressure *(Goulamoussene Biogeosciences)*. 
The cornerstone of tropical forests ecology is then to understand the underlyings of ecosystem response to disturbance, its mechanisms and drivers [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density *(e.g. PiponiotElife, RutishauserBiotropica)*. 
These structural parameters have then been sucessfully modeled and allowed to assess the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016]. 
However the response of tree species diversity remains unclear, also it is determinant of ecosystems productivity, stability and functioning [@Tilman2014, Liang2016Science] and would be most probably impacted by post-disturbance environmental changes [@BaralotoJAE.

@@

The functional diversity trajectories of plots based on the 8 leaf, stem and life history traits was examined through the Rao diversity.
For all undisturbed plots the functional diversity remained comparable to the reference values 5 years after disturbance (1989 inventories).
For all disturbed plots the functional trajectories followed a unimodal trajectory with a maximum positively correlated to disturbance intensity ($\rho_{spearman}=0.73$) (Figure \@ref(fig:DivFun)). 
The plot 7 from treatment 1 that displayed a constantly outlying diversity was removed from the graphical representation for better readability, see appendix for full graph.
Thirty years after disturbance all plots, whenever the initial disturbance intensity, regained diversity values similar to their initial value and to those of control plots.

@@

The functional redundancy was measured as the sum of weighted overlap among species in the communities.
Species functional space were defined in the main two-dimensional space of a PCA analysis based on all leaf and stem and maximum height traits (see appendix I for more details). 
Plots functoinal redundancy was followed in comparison to their initial value right after logging so that only communities dynamics are followed apart from the impact of logging itself.
While the functional redundancy of control plots remained stable along the 30 years it systematically decreased after disturbance (\@ref(fig:RedFun)).
The decrease of functional redundancy was comparable between all disturbed plots but started later for the most disturbed plots.
Plots recovery did not seemed correlated to the initial disturbance, for five out of nine disturbed plots the initial functional redundancy was had recovered while it was still decreasing or stabilized at lower values for the fou other plots.

```{r RedFun, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes.Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure"}
load("DB/Redundancy2")

par(mar=c(5,4,5,2))
RedundancyPlot(RedundancyTraj)
mtext("Functional Redundancy",side=3,adj=0,line=1.5)
mtext("Sum of abundance-weighted Trait Probability Density",side=3,adj=0,cex=0.8,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1)

@@

Besides the long-term alteration of functional redundancy, there was probably persistent compositional changes favoring disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and environmental changes, like in the soils nutrient cycling and compaction [@Olander2005].
These presistent changes highly question forest's resilience [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   #Pour avoir plusieurs lignes qui suivent. Commencer à 2 caractères
  Résumé de l'article.
keywords: [mot-clés, séparés par des virgules]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
---

@@

In tropical forest, ecological communities are contantly re-shaped by the natural disturbance events that changes the abiotic environment, through the fluxes of light, heat and water, and the biotic interaction and competitive pressure *(Goulamoussene Biogeosciences)*. 
The cornerstone of tropical forests ecology is then to understand the underlyings of ecosystem response to disturbance, its mechanisms and drivers [@White2001;@Chazdon2003a]. 
For now, this has been largely studied through structural parameters, rapid and convenient to measure, as aboveground biomass, tree height or stem density *(e.g. PiponiotElife, RutishauserBiotropica)*. 
These structural parameters have then been sucessfully modeled and allowed to assess the maintenance of ecosystems processes and services [@Denslow2000;@Blanc2009;@Rutishauser2016]. 
However the response of tree species diversity remains unclear, also it is determinant of ecosystems productivity, stability and functioning [@Tilman2014, Liang2016Science] and would be most probably impacted by post-disturbance environmental changes [@BaralotoJAE].

@@

The functional diversity trajectories of plots based on the 8 leaf, stem and life history traits was examined through the Rao diversity.
For all undisturbed plots the functional diversity remained comparable to the reference values 5 years after disturbance (1989 inventories).
For all disturbed plots the functional trajectories followed a unimodal trajectory with a maximum positively correlated to disturbance intensity ($\rho_{spearman}=0.73$) (Figure \@ref(fig:IDHplot)). 
The plot 7 from treatment 1 that displayed a constantly outlying diversity was removed from the graphical representation for better readability, see appendix for full graph.
Thirty years after disturbance all plots, whenever the initial disturbance intensity, regained diversity values similar to their initial value and to those of control plots.

@@

The functional redundancy was measured as the sum of weighted overlap among species in the communities.
Species functional space were defined in the main two-dimensional space of a PCA analysis based on all leaf and stem and maximum height traits (see appendix I for more details). 
Plots functional redundancy compared to the initial value after logging so that only communities dynamics are followed apart from the impact of logging itself.
Communities functional redundancy remained stable in control plots but after disturbance the redundancy trajectories were quite variable (See appendix I).
Globally after most intense disturbance (plots T2 and T3) communities redundancy decreased at first place and then increased to edge, recover or even exceed the initial value.
Considering the functional redundancy restricted to the functional space of the initial inventory, though, all disturbed plots followed similar decreasing humped-shaped trajectories (\@ref(fig:RedFun)).
The maximum redundancy loss observed was positively correlated with the disturbance intensity (XX spearman to be measured) and none of the disturbed plots had recovered the functional redundancy within their initial functional space.

```{r RedFun, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes.Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure"}
load("DB/Redundancy_restricted")

par(mar=c(5,4,5,2))
RedundancyPlot(RedundancyTraj_restricted)
mtext("Functional Redundancy",side=3,adj=0,line=1.5)
mtext("Restricted to initial functional space",side=3,adj=0,cex=0.8,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1)

@@

Besides the long-term alteration of functional redundancy, there was probably persistent compositional changes favoring disturbance resistant species [@Haddad2008], lianas or epiphytes [@Martin2013] and environmental changes, like in the soils nutrient cycling and compaction [@Olander2005].
These persistent changes highly question forest's resilience [@Chazdon2003a].
New conditions would not only be longer lasting but self-maintained as tied to disturbance regime [@Burslem2000].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

### 2018-05-11 WholePlotTrajectories.tex
Redundancy plot

The large areas covered with tropical forests worldwide hold crucial
economic, social and cultural value. They provide wood and multiple
non-timber forest products, shelter for a diversified fauna, regulate
the local climate, support the carbon, water and nutrient cycles, and
ensure cultural and human well-being. The simultaneous increase of
forests products demand and substantial climatic changes heightened the
pressure on the remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015} and threatened the maintenance
of communities structure, composition and functioning and their
underlying dynamics \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forests communities are shaped by a constant range of
disturbance that changes the abiotic environment, as the light, the heat
and water fluxes, and the biotic interaction and competitive pressure.
The cornerstone of tropical forests ecology is then to understand their
response to disturbance and its mechanisms
\citep{White2001, Chazdon2003a}. Forests response has been largely
studied through structural parameters rapid and convenient to measure,
as aboveground biomass, tree height or stem density. These structural
parameters have then been sucessfully modeled and allowed to assess the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009, Rutishauser2016}. However the response of
tree species diversity, which is determinant of ecosystems productivity,
stability and functioning \citep{Tilman2014} and would be most probably
impacted by post-disturbance environmental changes
\citep{CazzollaGatti2014}, remains unclear.

In the short-term diversity dynamics demonstrated negligible or even
positive impacts of disturbance on communities diversity, which have
been formalized by the intermediate disturbance hypothesis (IDH) stating
a maximized species diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. However, validations of
the IDH in the long term remain scarce and mainly rely on the analysis
of species richness that gives limited or misleading information on
forests recovery and functioning \citep{Martin2015, Chaudhary2016}. More
releveant monitoring would encompass communities composition, that is
crucial for conservation issues, and evenness, that reveals ecological
rules underlying communities' structure
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, the
functional approach accounting for species biological attributes would
be insightful as it reveals species fitness and ecosystems functioning
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect major functional traits related to species ecology and
performance were largely adopted through a relevant framework
\citep{Diaz2005, Villeger2008a}. The functional trait-based approach,
for example, highlighted in tropical rainforests the environmental
filters fostering disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. This
translated by shifts from ``conservative'' slow-growing species dealing
with scarce resources, to ``acquisitive'' fast-growing species with
rapid and efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. It was mirrored by shifts
in key functional traits related to resource acquisition (leaf area,
density and chlorophyll content, and stem specific gravity and bark
thickness), tree growth and reproduction life history traits (seed mass
and maximum height) \citep{Wright2004, Westoby2006a, Chave2009b}. A
proper monitoring of communities response should therefore encompass
taxonomic and functional diversity and composition measures to test the
validity of the IDH in the long term for tropical forests, and clarify
the resilience of communities evenness, composition, and also
functioning. The trajectories followed by all these facets would
highlight the role of deterministic processes, like competitive
exclusion or abiotic selection, and the communities' convergence
maintaining intrinsinc differences in diversity and composition,which is
as much insgights for future adaptive conservation strategies
\citep{Adler2007}.

Here we investigated over 30 years the response of 75 ha of forests
plots set up on a gradient of disturbance intensity, from 10 to 60\% of
ecosystem biomass removed. We made use of a large functional traits
database browsing major leaf, stem and seed traits and species maximum
height to draw the trajectories over time of communities taxonomic and
functional composition and diversity. Specifically, we (i) tested the
validity of the IDH in the long term for tropical hiperdiverse forest
and highlighted the ecological rules shaping their response to
disturbance, (ii) clarified the different facets of communities
resilience in terms of communities composition, diversity and
functioning (iii) questioned the completeness of communities recovery
given the altered functional redundancy.
\textgreater{}\textgreater{}revoir la fin

@@

The functional redundancy was measured as the overlap among species in
communities' functional space \citep{Carmona2016}. The samples of the
trait database were first located with a PCA analysis in a
two-dimensional functional space summarizing the 8 functional traits
considered. Then, multivariate kernel density estimator associated with
individual trees were summed to give the distribution of traits
probabilites of each species. For each community the trait probability
distributions of corresponding species were combined and weighted by
species abundance. Eventually communities functional redundancy was
measured as the sum of overlap between species weighted functional
density.

@@

The functional redundancy was measured as the sum of weighted overlap
among species in the communities. Species functional space were defined
in the main two-dimensional space of a PCA analysis based on all leaf
and stem and maximum height traits (see appendix I for more details).
Plots functoinal redundancy was followed in comparison to their initial
value right after logging so that only communities dynamics are followed
apart from the impact of logging itself. While the functional redundancy
of control plots remained stable along the 30 years it systematically
decreased after disturbance (\ref{fig:RedFun}). The decrease of
functional redundancy was comparable between all disturbed plots but
started later for the most disturbed plots. Plots recovery did not
seemed correlated to the initial disturbance, for five out of nine
disturbed plots the initial functional redundancy was had recovered
while it was still decreasing or stabilized at lower values for the fou
other plots.

@@

Besides the long-term alteration of functional redundancy, there was
probably persistent compositional changes favoring disturbance resistant
species \citep{Haddad2008}, lianas or epiphytes \citep{Martin2013} and
environmental changes, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These presistent changes highly question forest's
resilience \citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.

>>>>

The large areas covered with tropical forests worldwide hold crucial
economic, social and cultural value. They provide wood and multiple
non-timber forest products, shelter a diversified fauna, regulate the
local climate, support the carbon, water and nutrient cycles, and ensure
cultural and human well-being. The simultaneous increase of forests
products demand and substantial climatic changes currently heighten the
pressure on the remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}. It threatens the maintenance
of communities structure, composition and functioning and the underlying
dynamics in space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are contantly re-shaped by
the natural disturbance events that changes the abiotic environment,
through the fluxes of light, heat and water, and the biotic interaction
and competitive pressure \emph{(Goulamoussene Biogeosciences)}. The
cornerstone of tropical forests ecology is then to understand the
underlyings of ecosystem response to disturbance, its mechanisms and
drivers \citep{White2001, Chazdon2003a}. For now, this has been largely
studied through structural parameters, rapid and convenient to measure,
as aboveground biomass, tree height or stem density
\emph{(e.g.~PiponiotElife, RutishauserBiotropica)}. These structural
parameters have then been sucessfully modeled and allowed to assess the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009, Rutishauser2016}. However the response of
tree species diversity remains unclear, also it is determinant of
ecosystems productivity, stability and functioning
\citep[Liang2016Science]{Tilman2014} and would be most probably impacted
by post-disturbance environmental changes \citep{BaralotoJAE}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Validations of the IDH in
the long term, however, remain scarce and mainly rely on the analysis of
the rough species richness that gives limited or misleading information
on forests recovery and functioning \citep{Martin2015, Chaudhary2016}.
More releveant monitoring would encompass communities composition, that
is crucial for conservation issues, and diversity profiles, that reveals
ecological rules underlying communities' structure
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, the
functional approach accounting for differenes species biological
attributes would be insightful as it reveals the species role in the
post-disturbance trajectory of ecosystems functioning
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect major functional traits related to species ecology and
performance were largely adopted through an integrative framework
\citep{Diaz2005, Villeger2008a}. The functional trait-based approach,
for example, highlighted in tropical rainforests the environmental
filters fostering disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. In
disturbed tropical rainforests, this framework translates into the
fostering disturbance resistant species with rapid growth and efficient
resources acquisition \citep{Molino2001, Haddad2008} with a shift from
``conservative'' slow-growing species dealing with scarce resources that
dominate before disturbance, to ``acquisitive'' fast-growing species
with rapid and efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. It was mirrored by shifts
in key functional traits related to resource acquisition (leaf area,
density and chlorophyll content, and stem specific gravity and bark
thickness), tree growth and reproduction life history traits (seed mass
and maximum height) \citep{Wright2004, Westoby2006a, Chave2009b}.

A proper monitoring of communities response should therefore encompass
taxonomic and functional diversity and composition measures to test the
validity of the IDH in the long term for tropical forests, and clarify
the resilience of communities evenness, composition, and also
functioning. The trajectories followed by all these facets would
highlight the role of deterministic processes, like competitive
exclusion or abiotic selection, and the communities' convergence
maintaining intrinsinc differences in diversity and composition,which is
as much insgights for future adaptive conservation strategies
\citep{Adler2007}.

Here we investigated over 30 years the response of 75 ha of forests
plots set up on a gradient of disturbance intensity, from 10 to 60\% of
ecosystem biomass removed. We made use of a large functional traits
database browsing major leaf, stem and seed traits and species maximum
height to draw the trajectories over time of communities taxonomic and
functional composition and diversity. Specifically, we (i) tested the
validity of the IDH in the long term for tropical hyperdiverse forest
and highlighted the ecological rules shaping their response to
disturbance, (ii) clarified the different facets of communities
resilience in terms of communities composition, diversity and
functioning \emph{????(iii) questioned the completeness of communities
recovery given the altered functional redundancy.}

@@

The functional redundancy was measured as the overlap among species in
communities' functional space \citep{Carmona2016}. The samples of the
trait database were first located with a PCA analysis in a
two-dimensional functional space summarizing the 8 functional traits
considered. We choose to summarize species functional space through a
PCA because functional traits considered were already somehow redundant
and allows not to overweight the role of one trait. Then, multivariate
kernel density estimator associated with individual trees were summed to
give the distribution of traits probabilites of each species. For each
community the trait probability distributions of corresponding species
were combined and weighted by species abundance. Eventually communities
functional redundancy was measured as the sum of overlap between species
weighted functional density. Communities functional redundancy is
expressed as the average number of species that could be removed from
the community without reducing the functional space.

@@

The functional redundancy was measured as the sum of weighted overlap
among species in the communities. Species functional space were defined
in the main two-dimensional space of a PCA analysis based on all leaf
and stem and maximum height traits (see appendix I for more details).
Plots functional redundancy compared to the initial value after logging
so that only communities dynamics are followed apart from the impact of
logging itself. Communities functional redundancy remained stable in
control plots but after disturbance the redundancy trajectories were
quite variable (See appendix I). Globally after most intense disturbance
(plots T2 and T3) communities redundancy decreased at first place and
then increased to edge, recover or even exceed the initial value.
Considering the functional redundancy restricted to the functional space
of the initial inventory, though, all disturbed plots followed similar
decreasing humped-shaped trajectories (\ref{fig:RedFun}). The maximum
redundancy loss observed was positively correlated with the disturbance
intensity (XX spearman to be measured) and none of the disturbed plots
had recovered the functional redundancy within their initial functional
space.

@@

Besides the long-term alteration of functional redundancy, there was
probably persistent compositional changes favoring disturbance resistant
species \citep{Haddad2008}, lianas or epiphytes \citep{Martin2013} and
environmental changes, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These persistent changes highly question forest's
resilience \citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.
